---
title: "Step 1 Data Preparation"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Load packages and data

This chunk loads the required packages, checks that the Excel file exists, and reads the merged dataset into memory.

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(writexl)

data_path <- file.path("data", "merged_allthreedatasets.xlsx")

if (!file.exists(data_path)) {
  stop("Expected file not found: ", data_path)
}

merged_data <- read_xlsx(data_path)
merged_data
```

## Filter districts

Here we drop rows with unknown or excluded district codes to focus the analysis on valid districts.

```{r}
# Remove unknown and excluded district codes
filtered_data <- merged_data %>%
  filter(district_code != "Unk") %>%
  filter(!district_code %in% c("GT002", "GT101", "GT102", "WC011"))

filtered_data
```

## Column names of filtered data

This chunk lists the variable names available after filtering.

```{r}
colnames(filtered_data)
```

## Expand districts by years

We create every combination of district and years 2013–2023, then left-join back to the filtered data so missing district-year pairs are explicitly present.

```{r}
# 1. Get unique districts from filtered data
districts <- unique(filtered_data$district_code)

# 2. Create a vector of years
years <- 2013:2023

# 3. Cross–join districts and years
all_district_years <- expand.grid(
  district_code     = districts,
  year              = years,
  stringsAsFactors  = FALSE
)

# 4. Join back to filtered data
result <- all_district_years %>%
  left_join(filtered_data, by = c("district_code", "year"))

result
```






## Data coverage by year and variable

This chunk checks, for each variable and year, how many districts actually have data and what percent of the 52 districts that represents.

```{r}
coverage_by_year_var <- result %>%
  pivot_longer(
    cols      = -c(district_code, year),
    names_to  = "variable",
    values_to = "value"
  ) %>%
  mutate(has_data = !is.na(value)) %>%
  group_by(year, variable) %>%
  summarise(
    districts_with_data = sum(has_data),
    coverage_pct        = districts_with_data / 52 * 100,
    .groups             = "drop"
  ) %>%
  arrange(variable, year)

coverage_by_year_var
```

## Filter for high coverage

This chunk keeps only the variable-year combinations where at least 80% of districts reported data.

```{r}
coverage_filtered <- coverage_by_year_var %>%
  filter(coverage_pct >= 80)

coverage_filtered
```

## Count years with sufficient coverage per variable

This chunk counts how many years each variable meets the 80% coverage threshold.

```{r}
years_per_variable <- coverage_filtered %>%
  group_by(variable) %>%
  summarise(
    years_with_sufficient_coverage = n_distinct(year),
    .groups = "drop"
  ) %>%
  arrange(desc(years_with_sufficient_coverage))

years_per_variable
```

## How many variables have good coverage in at least 3 years?

This chunk counts variables that meet the 80% coverage threshold in three or more years.

```{r}
num_vars_ge3 <- years_per_variable %>%
  filter(years_with_sufficient_coverage >= 3) %>%
  tally() %>%
  pull(n)

cat("Number of variables with \u22653 years of sufficient coverage:", num_vars_ge3, "\\n")
```

## Which variables meet the 3+ years coverage threshold?

This chunk lists the variable names that have sufficient coverage (≥80%) in at least three years.

```{r}
variables_ge3_years <- years_per_variable %>%
  filter(years_with_sufficient_coverage >= 3) %>%
  pull(variable)

variables_ge3_years
```

## Keep only variables with 3+ years of good coverage

This chunk selects the district, year, and the variables that met the 3+ years coverage threshold.

```{r}
final_data <- result %>%
  dplyr::select(
    district_code,
    year,
    dplyr::all_of(variables_ge3_years)
  )

final_data
```

## Column names of the final dataset

This chunk lists the column names in `final_data`.

```{r}
colnames(final_data)
```

## Export the final dataset

This chunk saves the cleaned dataset to Excel in the `data` folder.

```{r}
output_path <- file.path("data", "step1_datapreparation_output.xlsx")
write_xlsx(final_data, output_path)
output_path
```

## Next steps
