---
title: "Step 2 Imputation (missForest)"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Load packages and Step 1 data

```{r}
library(readxl)
library(dplyr)
library(missForest)
library(mgcv)
library(zoo)
library(ggplot2)

# Path to the cleaned dataset from Step 1
data_path <- file.path("data", "step1_datapreparation_output.xlsx")

if (!file.exists(data_path)) {
  stop("Expected file not found: ", data_path, ". Run step1_datapreparation first.")
}

step1_data <- read_xlsx(data_path)
step1_data
```

```{r}
# Helper: clean names with janitor if available, otherwise fall back to base
clean_names_safe <- function(x) {
  if (requireNamespace("janitor", quietly = TRUE)) {
    return(janitor::make_clean_names(x))
  }
  message("Package 'janitor' not installed; using base::make.names().")
  make.names(x, unique = TRUE)
}
```

## Get South Africa district boundaries (GADM level 2)

This chunk downloads the South Africa GADM level-2 (district) boundaries, stores them under `data/spatial`, and returns an sf object `sa_gadm2_sf`.

```{r}
if (!dir.exists("data/spatial")) dir.create("data/spatial", recursive = TRUE)

library(geodata)
library(sf)

gadm_v <- geodata::gadm("ZAF", level = 2, path = "data/spatial")  # SpatVector
sa_gadm2_sf <- sf::st_as_sf(gadm_v)

sa_gadm2_sf
```

## Compute centroids and merge lon/lat into Step 1 data

```{r}
district_centroids <- sa_gadm2_sf %>%
  dplyr::rename(district_code = CC_2) %>%      # CC_2 holds the district code
  sf::st_centroid() %>%
  dplyr::mutate(
    lon = sf::st_coordinates(.)[, 1],
    lat = sf::st_coordinates(.)[, 2]
  ) %>%
  sf::st_drop_geometry() %>%
  dplyr::select(district_code, lon, lat)

step1_data_with_coords <- step1_data %>%
  dplyr::left_join(district_centroids, by = "district_code")

head(step1_data_with_coords)
```

```{r}


# Count records per district_code
step1_data_with_coords %>%
  dplyr::count(district_code, sort = TRUE)



#print(step1_data_with_coords)
```


### Check district/year coverage (2013 and 2023)

```{r}
required_years <- c(2013, 2023)
all_districts <- unique(district_centroids$district_code)

expected_pairs <- expand.grid(
  district_code = all_districts,
  year = required_years,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

observed_pairs <- step1_data_with_coords %>%
  dplyr::select(district_code, year) %>%
  dplyr::distinct()

missing_pairs <- expected_pairs %>%
  dplyr::anti_join(observed_pairs, by = c("district_code", "year"))

if (nrow(missing_pairs) == 0) {
  message("All district codes present for years 2013 and 2023.")
} else {
  message("Missing district/year combinations for 2013 or 2023:")
  print(missing_pairs)
}
```

## Column names of the merged data

```{r}
# Count records per district_code
step1_data_with_coords %>%
  dplyr::count(district_code, sort = TRUE)


```

## Impute missing values with GAM (spatio-temporal)

We fit a GAM for each variable (lon/lat/year smooth), impute missing values, then do a final linear interpolation over time within districts.

```{r}
# Variables with >=3 years of sufficient coverage
variables_ge3_years <- c(
  "Hospital beds per 10 000 target population (rescaled)",
  "Antenatal client initiated on ART rate",
  "Diarrhoea case fatality under 5 years rate",
  "Hospital beds per 10 000 target population",
  "Pneumonia case fatality under 5 years rate",
  "Severe acute malnutrition case fatality under 5 years rate",
  "ART Adult client viral load done (VLD)",
  "Adult ART Total",
  "Medical practitioners per 100 000 population",
  "No of households",
  "Percentage of households with access to improved sanitation (index)",
  "Pharmacists per 100 000 population",
  "Population.y",
  "Clients remaining on ART rate",
  "Delivery by Caesarean section rate (district hospitals)",
  "Proportion of health facilities with essential medicines",
  "Tracer items stock-out rate (fixed clinic/CHC/CDC)",
  "Antenatal client HIV 1st test positive",
  "Antenatal client known HIV positive but NOT on ART at 1st visit",
  "HIV test positive 15 years and older (excl ANC)",
  "Death under 5 years against live birth rate",
  "Access to services (Electricity)",
  "Access to services (Sewerage and sanitation)",
  "Access to services (Solid waste management)",
  "Access to services (Water)",
  "Acute malnutrition under five",
  "Adult literacy",
  "GDPR per capita",
  "Gini coefficient",
  "Human Development Index",
  "Immunisation rate",
  "Infant mortality",
  "Low Birth Weight",
  "Mean years of schooling",
  "Patients remaining on ART",
  "Patients starting ART treatment",
  "Under five mortality",
  "Expected years of schooling",
  "HIV test positive client 15 years and older rate (incl ANC)",
  "Maternal mortality"
)

# Build data for imputation: predictors (year, lon, lat) + targets
final_data_with_coords <- step1_data_with_coords %>%
  dplyr::select(
    district_code,
    year,
    lon,
    lat,
    dplyr::all_of(variables_ge3_years)
  )

# Clean names for modeling and keep a map back to originals
clean_name_map <- data.frame(
  orig = c("district_code", "year", "lon", "lat", variables_ge3_years),
  stringsAsFactors = FALSE
) %>%
  dplyr::mutate(clean = clean_names_safe(orig))

vars_to_impute <- setdiff(clean_name_map$clean, c("district_code", "year", "lon", "lat"))
if (length(vars_to_impute) == 0) stop("No variables available for imputation.")

gam_df <- final_data_with_coords %>%
  dplyr::rename_with(.fn = clean_names_safe) %>%
  dplyr::select(dplyr::all_of(clean_name_map$clean))

set.seed(123)
for (v in vars_to_impute) {
  message("Imputing with GAM: ", v)
  obs <- gam_df %>% dplyr::filter(!is.na(.data[[v]]))
  if (nrow(obs) < 10) {
    message("  • skipping ", v, "; only ", nrow(obs), " obs")
    next
  }
  k_year <- min(5, length(unique(obs$year)))
  k_spatial <- 10
  f_str <- sprintf("%s ~ te(lon, lat, year, k = c(%d,%d,%d))", v, k_spatial, k_spatial, k_year)
  fm <- as.formula(f_str)
  m <- mgcv::gam(fm, data = obs, method = "REML")
  preds <- predict(m, newdata = gam_df)
  gam_df[[v]] <- ifelse(is.na(gam_df[[v]]), preds, gam_df[[v]])
}

# Final temporal interpolation within districts
gam_df <- gam_df %>%
  dplyr::arrange(district_code, year) %>%
  dplyr::group_by(district_code) %>%
  dplyr::mutate(across(
    dplyr::all_of(vars_to_impute),
    ~ zoo::na.approx(.x, x = year, na.rm = FALSE, rule = 2)
  )) %>%
  dplyr::ungroup()

# Rename back to original column names
name_map <- stats::setNames(clean_name_map$orig, clean_name_map$clean)

final_imputed_gam <- gam_df %>%
  dplyr::rename_with(.cols = intersect(names(name_map), names(gam_df)), .fn = ~ name_map[.x])

head(final_imputed_gam)
```

## Check for remaining missing values
```{r}
colSums(is.na(final_imputed_gam))
```


```{r}
# Count records per district_code
final_imputed_gam %>%
  dplyr::count(district_code, sort = TRUE)
```


## Save imputed dataset
```{r}
output_path <- file.path("data", "step2_imputation_gam_output.xlsx")
write_xlsx(final_imputed_gam, output_path)
output_path
```

## Impute missing values with missForest (for comparison)

```{r}
mf_data <- final_data_with_coords %>%
  dplyr::select(year, lon, lat, dplyr::all_of(variables_ge3_years)) %>%
  as.data.frame()

set.seed(123)
mf_result <- missForest(
  xmis         = mf_data,
  maxiter      = 10,
  ntree        = 100,
  variablewise = FALSE,
  verbose      = TRUE
)

completed_mf <- mf_result$ximp

final_imputed_mf <- final_data_with_coords %>%
  dplyr::select(-dplyr::all_of(variables_ge3_years)) %>%
  dplyr::bind_cols(completed_mf %>% dplyr::select(dplyr::all_of(variables_ge3_years)))

# Save missForest version separately
output_path_mf <- file.path("data", "step2_imputation_missforest_output_mf.xlsx")
write_xlsx(final_imputed_mf, output_path_mf)
output_path_mf
```

## Flag spikes using robust z-score (MAD) by district

This chunk flags records where |z| > 3 using median/MAD within each district and variable.

```{r}
flag_mad_spikes <- function(df, dataset_label, out_dir) {
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  long_df <- df %>%
    pivot_longer(
      cols = -c(district_code, year),
      names_to = "variable",
      values_to = "value"
    ) %>%
    group_by(district_code, variable) %>%
    mutate(
      med = median(value, na.rm = TRUE),
      mad = stats::mad(value, constant = 1, na.rm = TRUE),
      z_mad = ifelse(mad == 0 | is.na(mad), NA_real_, (value - med) / mad),
      is_spike = !is.na(z_mad) & abs(z_mad) > 3
    ) %>%
    ungroup()

  spikes <- long_df %>%
    filter(is_spike) %>%
    arrange(district_code, variable, year)

  write.csv(
    spikes,
    file = file.path(out_dir, paste0("mad_spikes_", dataset_label, ".csv")),
    row.names = FALSE
  )

  spikes
}

spikes_gam <- flag_mad_spikes(final_imputed_gam, "final_imputed_gam", file.path("data", "checks"))
spikes_mf  <- flag_mad_spikes(final_imputed_mf,  "final_imputed_mf",  file.path("data", "checks"))

spikes_gam
spikes_mf

mad_spike_counts <- function(df) {
  df %>%
    pivot_longer(
      cols = -c(district_code, year),
      names_to = "variable",
      values_to = "value"
    ) %>%
    group_by(district_code, variable) %>%
    mutate(
      med = median(value, na.rm = TRUE),
      mad = stats::mad(value, constant = 1, na.rm = TRUE),
      z_mad = ifelse(mad == 0 | is.na(mad), NA_real_, (value - med) / mad),
      is_spike = !is.na(z_mad) & abs(z_mad) > 3
    ) %>%
    ungroup() %>%
    count(is_spike)
}

mad_spike_counts(final_imputed_gam)
mad_spike_counts(final_imputed_mf)

mad_spikes_by_variable <- function(df) {
  df %>%
    pivot_longer(
      cols = -c(district_code, year),
      names_to = "variable",
      values_to = "value"
    ) %>%
    group_by(district_code, variable) %>%
    mutate(
      med = median(value, na.rm = TRUE),
      mad = stats::mad(value, constant = 1, na.rm = TRUE),
      z_mad = ifelse(mad == 0 | is.na(mad), NA_real_, (value - med) / mad),
      is_spike = !is.na(z_mad) & abs(z_mad) > 3
    ) %>%
    ungroup() %>%
    filter(is_spike) %>%
    count(variable, sort = TRUE)
}

spike_vars_gam <- mad_spikes_by_variable(final_imputed_gam)
spike_vars_mf  <- mad_spikes_by_variable(final_imputed_mf)

spike_vars_gam
spike_vars_mf
```

```{r}
# Counts for GAM
spikes_gam %>%
  dplyr::count(is_spike)

# Counts for missForest
spikes_mf %>%
  dplyr::count(is_spike)
```




## Save line charts per variable and district

This chunk saves one PNG per variable for each imputed dataset. Each chart shows district lines across years.

```{r}
install.packages("ggplot2")

```

```{r}
library(ggplot2)

plot_district_lines <- function(df, vars, dataset_label, out_dir) {
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  for (v in vars) {
    p <- ggplot(df, aes(x = year, y = .data[[v]], group = district_code, color = district_code)) +
      geom_line(alpha = 0.6, linewidth = 0.5) +
      labs(
        title = paste0(dataset_label, ": ", v),
        x = "Year",
        y = v,
        color = "District"
      ) +
      theme_minimal() +
      theme(legend.position = "none")
    file_name <- paste0(
      dataset_label,
      "_",
      gsub("[^A-Za-z0-9_]+", "_", v),
      ".png"
    )
    ggsave(
      filename = file.path(out_dir, file_name),
      plot = p,
      width = 10,
      height = 6,
      dpi = 300
    )
  }
}

plot_district_lines(
  final_imputed_gam,
  variables_ge3_years,
  "final_imputed_gam",
  file.path("data", "checks")
)

plot_district_lines(
  final_imputed_mf,
  variables_ge3_years,
  "final_imputed_mf",
  file.path("data", "checks")
)
```

## Correlation screening and drop suggestions

This chunk flags highly correlated predictor pairs and suggests a variable to drop based on higher missingness in the pre-imputation data.

```{r}
find_high_corr <- function(df, vars, threshold = 0.85) {
  corr_mat <- cor(df[, vars], use = "pairwise.complete.obs")
  idx <- which(abs(corr_mat) >= threshold & abs(corr_mat) < 1, arr.ind = TRUE)
  if (nrow(idx) == 0) {
    return(data.frame())
  }
  data.frame(
    var1 = rownames(corr_mat)[idx[, 1]],
    var2 = colnames(corr_mat)[idx[, 2]],
    corr = corr_mat[idx],
    stringsAsFactors = FALSE
  ) %>%
    dplyr::mutate(pair = paste(pmin(var1, var2), pmax(var1, var2), sep = " | ")) %>%
    dplyr::distinct(pair, .keep_all = TRUE) %>%
    dplyr::select(-pair) %>%
    dplyr::arrange(dplyr::desc(abs(corr)))
}

suggest_drop <- function(pairs_df, missing_rate) {
  if (nrow(pairs_df) == 0) return(pairs_df)
  pairs_df %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      miss_var1 = missing_rate[var1],
      miss_var2 = missing_rate[var2],
      suggested_drop = dplyr::case_when(
        is.na(miss_var1) & is.na(miss_var2) ~ NA_character_,
        miss_var1 > miss_var2 ~ var1,
        miss_var2 > miss_var1 ~ var2,
        TRUE ~ NA_character_
      )
    ) %>%
    dplyr::ungroup()
}

# Missingness rates from pre-imputation data
missing_rate <- sapply(final_data_with_coords[, variables_ge3_years], function(x) mean(is.na(x)))

# High-correlation pairs for each imputed dataset
high_corr_gam <- find_high_corr(final_imputed_gam, variables_ge3_years, threshold = 0.85)
high_corr_mf  <- find_high_corr(final_imputed_mf,  variables_ge3_years, threshold = 0.85)

high_corr_gam
high_corr_mf

# Drop suggestions based on higher pre-imputation missingness
drop_suggestions_gam <- suggest_drop(high_corr_gam, missing_rate)
drop_suggestions_mf  <- suggest_drop(high_corr_mf,  missing_rate)

drop_suggestions_gam
drop_suggestions_mf

# Save drop suggestions (without suggested_drop column)
dir.create(file.path("data", "checks"), recursive = TRUE, showWarnings = FALSE)

drop_suggestions_gam %>%
  dplyr::select(-suggested_drop) %>%
  write.csv(
    file = file.path("data", "checks", "drop_suggestions_gam.csv"),
    row.names = FALSE
  )

drop_suggestions_mf %>%
  dplyr::select(-suggested_drop) %>%
  write.csv(
    file = file.path("data", "checks", "drop_suggestions_mf.csv"),
    row.names = FALSE
  )
```

## Compare GAM vs missForest imputations

```{r}
common_cols <- intersect(names(final_imputed_gam), names(final_imputed_mf))
num_cols <- common_cols[sapply(final_imputed_gam[common_cols], is.numeric)]

diff_counts <- sapply(num_cols, function(col) {
  sum(abs(final_imputed_gam[[col]] - final_imputed_mf[[col]]) > 1e-8, na.rm = TRUE)
})

total_diff <- sum(diff_counts)

message("Total numeric cells differing between GAM and missForest: ", total_diff)
print(diff_counts)
```

## Drop selected columns from both imputed datasets

```{r}
cols_to_drop <- c(
  "Access to services (Solid waste management)",
  "Access to services (Electricity)",
  "Antenatal client known HIV positive but NOT on ART at 1st visit",
  "ART Adult client viral load done (VLD)"
)

final_imputed_gam <- final_imputed_gam %>%
  dplyr::select(-dplyr::all_of(cols_to_drop))

final_imputed_mf <- final_imputed_mf %>%
  dplyr::select(-dplyr::all_of(cols_to_drop))

final_imputed_gam
final_imputed_mf
```

## Re-save outputs after dropping columns

```{r}
output_path <- file.path("data", "step2_imputation_gam_output.xlsx")
write_xlsx(final_imputed_gam, output_path)
output_path

output_path_mf <- file.path("data", "step2_imputation_missforest_output_mf.xlsx")
write_xlsx(final_imputed_mf, output_path_mf)
output_path_mf
```
