---
title: "Step 3 Clean AMR Data"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Load AMR CSV

```{r}
library(readr)
library(dplyr)
library(readxl)
library(writexl)

amr_path <- file.path("data", "sa_main_amr.csv")

if (!file.exists(amr_path)) {
  stop("Expected file not found: ", amr_path)
}

amr_data <- read_csv(amr_path, show_col_types = FALSE)
amr_data
```

## Load facility bug-drug yearly summary

```{r}
facility_summary_path <- file.path("data", "Facility_BugDrug_YearlySummary.xlsx")

if (!file.exists(facility_summary_path)) {
  stop("Expected file not found: ", facility_summary_path)
}

facility_summary <- read_excel(facility_summary_path)
facility_summary
```

## Load hospital tier data

```{r}
hospital_tier_path <- file.path("data", "Hospital_tier_final.xlsx")

if (!file.exists(hospital_tier_path)) {
  stop("Expected file not found: ", hospital_tier_path)
}

hospital_tier_data <- read_excel(hospital_tier_path)
hospital_tier_data
```

## Normalize drug susceptibility columns

This chunk standardizes drug result values to `Resistant` or `Sensitive` (case-insensitive) and sets anything else to `NA`.

```{r}
drug_cols <- c(
  "meropenem", "imipenem", "doripenem",
  "cefotaxime__ceftriaxone", "ceftazidime",
  "cloxacillin", "ertapenem",
  "gentamicin", "tobramycin", "amikacin", "netilmicin", "gentamicin___high_level_resistan",
  "vancomycin", "teicoplanin"
)

amr_data_normalized <- amr_data %>%
  mutate(across(
    all_of(drug_cols),
    ~ case_when(
      toupper(as.character(.)) %in% c("RESISTANT", "INTERMEDIATE") ~ "Resistant",
      toupper(as.character(.)) %in% c("SENSITIVE", "SENSITIVE DOSE DEPENDANT") ~ "Sensitive",
      TRUE ~ NA_character_
    )
  ))

amr_data_normalized
```

## Derive organism-specific resistance flags

```{r}
amr_data_flags <- amr_data_normalized %>%
  mutate(
    abaumanii_carbapenem = case_when(
      eskape != "A. baumannii" ~ NA_character_,
      rowSums(across(c(meropenem, imipenem, doripenem),
                     ~ . == "Resistant"), na.rm = TRUE) > 0 ~ "Resistant",
      rowSums(across(c(meropenem, imipenem, doripenem),
                     ~ is.na(.)), na.rm = TRUE) == 3 ~ NA_character_,
      TRUE ~ "Sensitive"
    ),

    ecoli_3gc = case_when(
      eskape != "E. coli" ~ NA_character_,
      rowSums(across(c(cefotaxime__ceftriaxone, ceftazidime),
                     ~ . == "Resistant"), na.rm = TRUE) > 0 ~ "Resistant",
      rowSums(across(c(cefotaxime__ceftriaxone, ceftazidime),
                     ~ is.na(.)), na.rm = TRUE) == 2 ~ NA_character_,
      TRUE ~ "Sensitive"
    ),

    saureus_penicillinase = case_when(
      eskape != "S. aureus" ~ NA_character_,
      cloxacillin == "Resistant" ~ "Resistant",
      is.na(cloxacillin) ~ NA_character_,
      TRUE ~ "Sensitive"
    ),

    kpneumoniae_carbapenem = case_when(
      eskape != "K. pneumoniae" ~ NA_character_,
      rowSums(across(c(meropenem, imipenem, ertapenem, doripenem),
                     ~ . == "Resistant"), na.rm = TRUE) > 0 ~ "Resistant",
      rowSums(across(c(meropenem, imipenem, ertapenem, doripenem),
                     ~ is.na(.)), na.rm = TRUE) == 4 ~ NA_character_,
      TRUE ~ "Sensitive"
    ),

    paeruginosa_aminoglycoside = case_when(
      eskape != "P. aeruginosa" ~ NA_character_,
      rowSums(across(c(gentamicin, tobramycin, amikacin, netilmicin,
                       gentamicin___high_level_resistan),
                     ~ . == "Resistant"), na.rm = TRUE) > 0 ~ "Resistant",
      rowSums(across(c(gentamicin, tobramycin, amikacin, netilmicin,
                       gentamicin___high_level_resistan),
                     ~ is.na(.)), na.rm = TRUE) == 5 ~ NA_character_,
      TRUE ~ "Sensitive"
    ),

    enterobacter_3gc = case_when(
      eskape != "Enterobacter spp" ~ NA_character_,
      rowSums(across(c(cefotaxime__ceftriaxone, ceftazidime),
                     ~ . == "Resistant"), na.rm = TRUE) > 0 ~ "Resistant",
      rowSums(across(c(cefotaxime__ceftriaxone, ceftazidime),
                     ~ is.na(.)), na.rm = TRUE) == 2 ~ NA_character_,
      TRUE ~ "Sensitive"
    ),

    efaecium_glycopeptide = case_when(
      eskape != "E. faecium" ~ NA_character_,
      rowSums(across(c(vancomycin, teicoplanin),
                     ~ . == "Resistant"), na.rm = TRUE) > 0 ~ "Resistant",
      rowSums(across(c(vancomycin, teicoplanin),
                     ~ is.na(.)), na.rm = TRUE) == 2 ~ NA_character_,
      TRUE ~ "Sensitive"
    ),

    efaecalis_glycopeptide = case_when(
      eskape != "E. faecalis" ~ NA_character_,
      rowSums(across(c(vancomycin, teicoplanin),
                     ~ . == "Resistant"), na.rm = TRUE) > 0 ~ "Resistant",
      rowSums(across(c(vancomycin, teicoplanin),
                     ~ is.na(.)), na.rm = TRUE) == 2 ~ NA_character_,
      TRUE ~ "Sensitive"
    )
  )

amr_data_flags
```

## Select final columns

```{r}
cols_to_keep <- c(
  "unique_patient_id",
  "taken_date",
  "organism_name",
  "episode_no",
  "folder_no",
  "gender",
  "date_of_birth",
  "facility_name",
  "facility_latitude",
  "facility_longitude",
  "province_name",
  "district_name",
  "ward_name",
  "ward_description",
  "referring_lab_name",
  "testing_lab_name",
  "lab_latitude",
  "lab_longitude",
  "registered_date",
  "tested_date",
  "reviewed_date",
  "eskape",
  "organism",
  "date_taken",
  "year_taken",
  "month",
  "abaumanii_carbapenem",
  "ecoli_3gc",
  "saureus_penicillinase",
  "kpneumoniae_carbapenem",
  "paeruginosa_aminoglycoside",
  "enterobacter_3gc",
  "efaecium_glycopeptide",
  "efaecalis_glycopeptide"
)

amr_data_final <- amr_data_flags %>%
  select(all_of(cols_to_keep))

amr_data_final
```

## Add hospital tier

```{r}
amr_data_final <- amr_data_final %>%
  left_join(
    hospital_tier_data %>% select(facility_name, hospital_tier),
    by = "facility_name"
  )

# Verify the join
amr_data_final %>%
  select(facility_name, hospital_tier) %>%
  distinct() %>%
  head()

head(amr_data_final)
```

## Column names

```{r}
colnames(amr_data_final)
```

## Districts in facility summary
```{r}
distinct_districts <- facility_summary %>%
  distinct(district_name, district_code)
distinct_districts

```

## Add district codes to AMR data

```{r}
amr_data_final <- amr_data_final %>%
  left_join(
    distinct_districts,
    by = "district_name"
  )

head(amr_data_final)
```

## Drop rows without a district code
```{r}
amr_data_final <- amr_data_final %>%
  filter(!is.na(district_code) & trimws(district_code) != "")
```


## Save cleaned AMR data
```{r}
print(colnames(amr_data_final))
```

## Rows missing district code
```{r}
amr_data_final %>%
  filter(is.na(district_code) | trimws(district_code) == "") %>%
  print(n = Inf)
```




## Drop rows without a district code
```{r}
amr_data_final <- amr_data_final %>%
  filter(!is.na(district_code) & trimws(district_code) != "")
  
```


## Save cleaned AMR data to Excel
```{r}
output_path <- file.path("data", "step3_cleanedAMRdata_output.xlsx")
write_xlsx(amr_data_final, output_path)
output_path
```